Well-quasi-orderings on word languages

### 2024-12-16 src/infixes.tex
Restructure
Ordinals

@@

\begin{lemma}
    \label{thue-morse-ordinal:lemma}
    The \kl{maximal order type} of $\LMorse$ is $\omega$,
    the \kl{ordinal height} of $\LMorse$ is $\omega$,
    the \kl{ordinal width} of $\LMorse$ is $\omega$.
\end{lemma}
\begin{proof}
    Let us prove that these are upper bounds for the \kl{ordinal invariants} of
    $L$. For the \kl{ordinal height}, it is true for any language $\LMorse$.
    For the \kl{maximal order type}, we remark that
    the maximal length of a \kl{bad sequence} is determined by its first element,
    hence that it is at most $\omega$.
    Finally, because the \kl{ordinal width} is at most the \kl{maximal order type},
    we conclude that the \kl{ordinal width} is also at most $\omega$.

    Now, let us prove that these bounds are tight. It is clear that
    $\oHeight{\LMorse} = \omega$: given any number $n \in \Nat$, one can construct a
    \kl{decreasing sequence} of words in $L$ of length $n$, for instance by
    considering the first $n$ prefixes of the \kl{Thue-Morse sequence} by
    decreasing size.
    Let us now prove that $\oWidth{\LMorse} = \omega$. Assume by contradiction that
    $\oWidth{\LMorse}$ is finite. Then, $L$ can be written as a finite union of
    \kl{chains} for the \kl{infix relation}, and in particular, $L$ is
    \kl(language){bounded}, which is absurd by \cref{thue-morse:lemma}.
    Finally, because the \kl{ordinal width} is at most the \kl{maximal order
    type}, we conclude that the \kl{maximal order type} of $\LMorse$ is also $\omega$.
\end{proof}

>>>>

\section{Infixes and Downwards Closed Languages}

@@

\begin{lemma}
    \label{thue-morse-ordinal:lemma}
    The \kl{maximal order type} of $\LMorse$ is $\omegaOrd$,
    the \kl{ordinal height} of $\LMorse$ is $\omegaOrd$,
    the \kl{ordinal width} of $\LMorse$ is $\omegaOrd$.
\end{lemma}
\begin{proof}
    Let us prove that these are upper bounds for the \kl{ordinal invariants} of
    $L$. For the \kl{ordinal height}, it is true for any language $\LMorse$.
    For the \kl{maximal order type}, we remark that
    the maximal length of a \kl{bad sequence} is determined by its first element,
    hence that it is at most $\omegaOrd$.
    Finally, because the \kl{ordinal width} is at most the \kl{maximal order type},
    we conclude that the \kl{ordinal width} is also at most $\omegaOrd$.

    Now, let us prove that these bounds are tight. It is clear that
    $\oHeight{\LMorse} = \omegaOrd$: given any number $n \in \Nat$, one can construct a
    \kl{decreasing sequence} of words in $L$ of length $n$, for instance by
    considering the first $n$ prefixes of the \kl{Thue-Morse sequence} by
    decreasing size.
    Let us now prove that $\oWidth{\LMorse} = \omegaOrd$. Assume by contradiction that
    $\oWidth{\LMorse}$ is finite. Then, $L$ can be written as a finite union of
    \kl{chains} for the \kl{infix relation}, and in particular, $L$ is
    \kl(language){bounded}, which is absurd by \cref{thue-morse:lemma}.
    Finally, because the \kl{ordinal width} is at most the \kl{maximal order
    type}, we conclude that the \kl{maximal order type} of $\LMorse$ is also $\omegaOrd$.
\end{proof}

### 2024-12-16 src/introduction.tex
Ordinals

Note that \kl{chains} are the simplest possible \kl{well-quasi-ordered} sets
(they are \kl{totally ordered} and \kl{well-founded}) and it is known that finite unions
of \kl{well-quasi-ordered} sets are \kl{well-quasi-ordered}. As a consequence,
the above characterization states that only the simplest possible
\kl{well-quasi-ordered} sets are \kl{well-quasi-ordered} by the \kl{prefix
relation}. Let us furthermore highlight that this characterization holds
without any restriction on the decidability of the language $L$ itself.\footnote{But
heavily relies on the assumption that $\Sigma$ is finite} This allows us to
derive the following bounds on the \kl{ordinal invariants} of such
\kl{well-quasi-ordered} languages, which can be interpreted in two dual ways:
first, these languages are extremely simple, which means that one could have
hoped for using directly another well-quasi-ordered set without resorting to
finite words; second, any time one encounters a \kl{well-quasi-ordered}
language, it is going to provide relatively efficient algorithms.
{
\renewcommand{\labelenumi}{R\arabic{enumi}}
\begin{enumerate}
    \setcounter{enumi}{2}
    \item All languages $L \subseteq \Sigma^*$ are 
        \kl{well-quasi-ordered} by the \kl{prefix relation} 
        (resp. \kl{suffix relation}) have an \kl{ordinal height} of at most $\omega$,
        a finite \kl{ordinal width}, and a \kl{maximal order type} strictly
        below $\omega^2$.
        \hfill (\cref{prefixes-ordinal-invariants:cor} p.\pageref{prefixes-ordinal-invariants:cor})
\end{enumerate}
}

\AP The straightforward generalization of the results for the \kl{prefix
relation} and the \kl{suffix relation} to the \kl{infix relation} is not
possible. Indeed, it follows from Kuske's result that $\Sigma^*$ equipped with
the \kl{subword relation} can be embedded into $\set{a,b}^*$ with the \kl{infix
relation}~\cite[Lemma 5.1]{DBLP:journals/ita/Kuske06}. This implies that there
are \kl{well-quasi-ordered} languages for the \kl{infix relation} that have
very large \kl{ordinal invariants}: for instance, the \kl{maximal order type}
of the \kl{subword relation} is $\omega^{\omega^{\card{\Sigma} - 1}}$, which
equals its \kl{ordinal width} \cite[Corollary 3.9, Theorem 4.21]{DZSCSC20}.
However, not all is lost if we restrict ourselves to \kl{bounded languages},
i.e., languages that are included in some $w_1^* \cdots w_k^*$ for some finite
choice of words $w_1, \ldots, w_k$. We are able to show the following:

@@

{
\renewcommand{\labelenumi}{R\arabic{enumi}}
\begin{enumerate}
    \setcounter{enumi}{5}
    \item 
        Let $L$ be a language recognized by an \kl{amalgamation system}.
        If $L$ is \kl{well-quasi-ordered} by the \kl{infix relation},
        then $L$ is a \kl{bounded language}.

>>>>

Note that \kl{chains} are the simplest possible \kl{well-quasi-ordered} sets
(they are \kl{totally ordered} and \kl{well-founded}) and it is known that finite unions
of \kl{well-quasi-ordered} sets are \kl{well-quasi-ordered}. As a consequence,
the above characterization states that only the simplest possible
\kl{well-quasi-ordered} sets are \kl{well-quasi-ordered} by the \kl{prefix
relation}. Let us furthermore highlight that this characterization holds
without any restriction on the decidability of the language $L$ itself.\footnote{But
heavily relies on the assumption that $\Sigma$ is finite} This allows us to
derive the following bounds on the \kl{ordinal invariants} of such
\kl{well-quasi-ordered} languages, which can be interpreted in two dual ways:
first, these languages are extremely simple, which means that one could have
hoped for using directly another well-quasi-ordered set without resorting to
finite words; second, any time one encounters a \kl{well-quasi-ordered}
language, it is going to provide relatively efficient algorithms.
{
\renewcommand{\labelenumi}{R\arabic{enumi}}
\begin{enumerate}
    \setcounter{enumi}{2}
    \item All languages $L \subseteq \Sigma^*$ are 
        \kl{well-quasi-ordered} by the \kl{prefix relation} 
        (resp. \kl{suffix relation}) have an \kl{ordinal height} of at most $\omegaOrd$,
        a finite \kl{ordinal width}, and a \kl{maximal order type} strictly
        below $\omegaOrd^2$.
        \hfill (\cref{prefixes-ordinal-invariants:cor} p.\pageref{prefixes-ordinal-invariants:cor})
\end{enumerate}
}

\AP The straightforward generalization of the results for the \kl{prefix
relation} and the \kl{suffix relation} to the \kl{infix relation} is not
possible. Indeed, it follows from Kuske's result that $\Sigma^*$ equipped with
the \kl{subword relation} can be embedded into $\set{a,b}^*$ with the \kl{infix
relation}~\cite[Lemma 5.1]{DBLP:journals/ita/Kuske06}. This implies that there
are \kl{well-quasi-ordered} languages for the \kl{infix relation} that have
very large \kl{ordinal invariants}: for instance, the \kl{maximal order type}
of the \kl{subword relation} is $\omegaOrd^{\omegaOrd^{\card{\Sigma} - 1}}$, which
equals its \kl{ordinal width} \cite[Corollary 3.9, Theorem 4.21]{DZSCSC20}.
However, not all is lost if we restrict ourselves to \kl{bounded languages},
i.e., languages that are included in some $w_1^* \cdots w_k^*$ for some finite
choice of words $w_1, \ldots, w_k$. We are able to show the following:

@@

{
\renewcommand{\labelenumi}{R\arabic{enumi}}
\begin{enumerate}
    \setcounter{enumi}{6}
    \item 
        Let $L$ be a language recognized by an \kl{amalgamation system}.
        If $L$ is \kl{well-quasi-ordered} by the \kl{infix relation},
        then $L$ is a \kl{bounded language}.

### 2024-12-16 src/prelims.tex
Ordinals

\paragraph*{Ordinal Invariants.} \label{ordinal-invariants:subsec}
An \intro{ordinal} is a \kl{well-founded} \kl{totally ordered}
set. We use $\alpha, \beta, \gamma$ to denote ordinals, and use $\omega$ to
denote the first infinite ordinal, i.e., the set of natural numbers with the
usual ordering. We assume superficial familiarity with ordinal arithmetic, and
refer to the books of Kunen \cite{KUNEN80} and Krivine~\cite[Chapter
II]{KRIVINE71} for a detailed introduction to this domain. Given a tree $T$
whose branches are all finite we can define an ordinal $\alpha_T$ inductively
as follows: if $T$ is a leaf then $\alpha_T = 0$, if $T$ has children
$\seqof{T_i}$ then $\alpha_T = \sup \setof{\alpha_{T_i} + 1}{i \in \Nat}$. We
say that $\alpha_T$ is the \emph{rank} of $T$.

>>>>

\paragraph*{Ordinal Invariants.} \label{ordinal-invariants:subsec}
An \intro{ordinal} is a \kl{well-founded} \kl{totally ordered}
set. We use $\alpha, \beta, \gamma$ to denote ordinals, and use $\intro*\omegaOrd$ to
denote the first infinite \kl{ordinal}, i.e., the set of natural numbers with the
usual ordering. We assume superficial familiarity with ordinal arithmetic, and
refer to the books of Kunen \cite{KUNEN80} and Krivine~\cite[Chapter
II]{KRIVINE71} for a detailed introduction to this domain. Given a tree $T$
whose branches are all finite we can define an \kl{ordinal} $\alpha_T$ inductively
as follows: if $T$ is a leaf then $\alpha_T = 0$, if $T$ has children
$\seqof{T_i}$ then $\alpha_T = \sup \setof{\alpha_{T_i} + 1}{i \in \Nat}$. We
say that $\alpha_T$ is the \emph{rank} of $T$.
